\Large \textbf{ Phenotype and gene ontology enrichment as guides for disease modeling in \cel{} }

### 2018-02-06 pea_paper_docs/tex/pea.tex

\usepackage[utf8x]{inputenc}
\usepackage{nameref}
\usepackage[right]{lineno}
\usepackage{amsmath}
\usepackage{booktabs}
\usepackage[numbers,super]{natbib}
\usepackage{changepage}

@@

@@

\author[1,2]{David Angeles-Albores}
\author[1]{Raymond Y. N. Lee}
\author[1]{Juancarlos Chan}
\author[1,2,*]{Paul W. Sterberg}
\affil[1]{Division of Biology and Biological Engineering, Caltech,
Pasadena, CA, 91125, USA}
\affil[2]{Howard Hughes Medical Institute, Caltech, Pasadena, CA, 91125, USA}
\renewcommand\Affilfont{\itshape\small{}}

@@

\textbf{1} Department of Biology and Biological Engineering,
and Howard Hughes Medical Institute, Caltech, Pasadena, CA, 91125, USA\\
\textbf{*} Corresponding author. Contact:pws@caltech.edu

\section*{Abstract}
\textbf{
Genome-wide experiments have the capacity to generate massive amounts of unbiased
data about an organism. To interpret these data, dimensionality reduction
techniques are required. One approach is to annotate genes using unified,
unambiguous vocabularies and to test experimental datasets for term enrichment
using probabilistic methods.
Although gene, phenotype and anatomy ontologies exist for \cel{}, no one
software offers enrichment analyses of all three ontologies.
Here, we present the WormBase Enrichment Suite, which offers users the
ability to test all nematode ontologies simultaneously. We show that
the WormBase Enrichment Suite provides valuable insight into different
biological problems, and that phenotype enrichment analysis (PEA) can
help researchers identify disease phenologs, phenotypes that are homologous across
species, which can inform disease modeling in \cel{}. The WormBase Enrichment Suite
analysis can also shed light on RNA-seq datasets by showing what
molecular functions are enriched, which phenotypes these functions are implicated
in and what tissues are overrepresented in the dataset. Finally, we explore the
phenotype-anatomy relationship, showing that a small subset of highly specific
tissues are disproportionately likely to cause an Egl phenotype, but inferring
tissue expression from an egg-laying (Egl) phenotype is limited to the largest tissues.
}

@@

\section*{Introduction}
The last decade has seen an explosion of techniques capable of genome-wide
measurements. Some examples of genome-wide tools include
RNA-seq~\cite{Mortazavi2008} to measure gene expression or
CHIP-seq~\cite{Johnson2007} to measure protein binding to chromatin. These tools
are capable of generating large quantities of data. Understanding these data,
and generating hypotheses from them remains challenging. A common approach used
to understand these datasets is to reduce the dimensionality of the data via
enrichment analyses of ontologies~\cite{TheGeneOntologyConsortium2000a}, which
helps researchers understand what terms are overrepresented beyond random
levels. By analyzing overrepresented terms in aggregate, researchers can better
understand what biological processes were most affected in a given experiment,
and form hypotheses about what is happening~\cite{Rhee2008}. This approach is
limited by what ontologies can be tested for enrichment. The best-known ontologies
for biological research are the Gene Ontologies (GO), which provide a controlled
language to describe the biological, cellular and molecular aspects and functions of
genes~\cite{TheGeneOntologyConsortium2000a}. In \cel{}, gene, tissue and
phenotype ontologies exist with which to describe \cel{} anatomy and phenotypes
respectively~\cite{Schindelman2011,Lee2003}. These latter ontologies are developed by
professional curators at WormBase, which is a repository of \cel{}
data~\cite{Howe2016}. However, enrichment tools currently only exist for gene and tissue
ontologies in the community today (see for
example~\cite{Chikina2009,Mi2013,Angeles-Albores2016}). Another limitation is
that tissue enrichment testing is not offered on the same websites as GO
enrichment testing, which requires users to test their data on different
websites that may or may not use different methodologies to detect enrichment.

Another way to use enrichment tools is for evolutionary comparison purposes.
In molecular biology it is often useful to know when a gene is homologous
between two species---that is to say, common by descent---because knowledge of
homology often brings with it knowledge of function. Indeed, many
important gene regulatory networks (GRNs) are conserved between organisms as
highly diverged as nematodes and humans (for example, see~\cite{Sternberg1998}).
While genes and GRNs may be conserved between species, their outputs often differ.
For example, the gene Pax6 (Eyeless in \emph{Drosophila melanogaster})
is involved in eye formation in humans and fruit-flies~\cite{Quiring1994}. Although
nematodes have conserved this gene, they do not have eyes~\cite{Zhang1995,Chisholm1995}.
The concept of a phenolog has been put forward to explain relationships between
phenotypes that have the same underlying genetic
regulatory network~\cite{McGary2010,Lehner2013}.
Formally, two phenotypes are phenologs of each other if the orthologs of the
genes that cause a phenotype in an organism cause a second phenotype in another.

To study a clinically relevant disease in a non-human, an appropriate model has
to be established. A straightforward method towards establishing a disease model
in \cel{} is to link a disease to a causal gene, then to identify the homologous
gene in \cel{} and then to study the function of the genetic homolog to
extrapolate back to humans. However, this method relies on the existence of
known disease genes and requires that the homolog have a phenotype that can be
reliably identified and studied. A fundamentally different way to establish a
disease model in \cel{} would be to identify the phenologs of the disease to be
studied in \cel{} by identifying disease-associated human genes in an unbiased
manner through genome-wide association studies (GWAS) and identified candidate
homolog genes in \cel{}. The orthologs can be used to identify \cel{} disease
phenologs, which can in turn be used as the basis for screens to identify genes
that are associated with that phenolog. Approaches similar to this have been
successfully used in the past to make non-obvious links between phenotypes in
different species~\cite{McGary2010}.

The concept of a phenolog can also be useful when applied within a species. In
\cel{}, not all phenotypes are equally easy to study. Although genome-wide
measurements can help elucidate the genetic network underlying a phenotype,
devising screens to test which genes are functionally important can be
difficult. A common strategy to study phenotypes that are difficult to screen is
to select an easier-to-screen phenolog, and to test positive hits for the true
phenotype of interest afterwards. For example, candidate genes to extend the
\cel{} lifespan can be first screened for using heat shock survival genes
involved in \cel{} aging sensitivity can be identified using stress
assays~\cite{Kim2007a,Mehta2009}. Currently, selection of screening phenotypes
is performed based on researcher experience. By formalizing phenotype enrichment
analysis as a tool with which to analyze gene sets, researchers should be able
to formally establish phenologs, which has consequences for screen design. %
eating defects in an ablated strain

An additional problem with genome-wide queries of \cel{} states (be they
developmental, such as L1, L2, dauer; behavioral states such as awake versus
asleep; or other) is that they do not always have a straightforward
interpretation in terms of phenotypes. In these situations, researchers must
rely on intuition to select a phenotype for which to screen. As a result, many
hits may go unexplored that would prove fruitful. The question of how to design
a screen that is maximally informative is an important question that has so far
not been addressed within this community.

To facilitate understanding of large datasets, we have developed the WormBase
Enrichment Suite (URL:\
\url{http://www.wormbase.org/tools/enrichment/tea/tea.cgi}) that allows users to
rapidly perform phenotype, tissue and gene ontology enrichment analyses (PEA,
TEA and GEA respectively) on curated \cel{} ontologies using the same
methodology for each one. As examples of use cases, we applied our tools towards
the unbiased discovery of phenologs of multigenic, complex diseases including
systemic lupus erythematosus, obesity and obesity-related traits, and rheumatoid
arthritis by using genes associated with these diseases via genome-wide
association studies. We illustrate the utility of the complete enrichment suite
for finding new relationships in complex data by analyzing a ciliary neuron
transcriptome~\cite{Wang2015}. Finally, we show that the dictionaries generated
for these enrichment analyses can help elucidate the contributions of specific
tissues to specific phenotypes.

@@

Enrichment analyses are performed as described in~\cite{Angeles-Albores2016}.
Briefly, a static dictionary is constructed for each ontology using the most
recent WormBase release annotation information. This dictionary is generated
using a trimming algorithm described in~\cite{Angeles-Albores2016} to remove
terms that are too general and uninformative as well as terms that are very
specific but do not have many annotations. When the user provides a set of genes
to test against, we use a hypergeometric model to test whether a given term
within an ontology is occurring at a greater frequency than expected by random
chance.

\subsection*{Human disease phenolog identification}
We used the GWAS EBI-NHGRI catalog~\cite{MacArthur2016} to extract information
on all genome-wide association studies deposited there. We only selected traits
that had $>300$ associated genes. We identified 24 traits that met our criteria.
Next, we used DIOPT~\cite{Hu2011} to identify candidate orthologs for the genes
associated with these traits. Briefly, DIOPT combines a large number of methods
for identifying orthologs and returns homolog candidates associated with a
compound score. Depending on the score, orthologs can be considered `high',
`moderate' or `low' rank, reflecting confidence in the homology. Many-to-one and
one-to-many homology relationships are allowed in DIOPT, reflecting a mixture of
uncertainty and family expansion/reduction. For our study, we only accepted
homolog candidates with `high' or `moderate' scores and we did not insist on a
one-to-one relationship between genes.

\begin{figure}[htbp]
  \renewcommand{\familydefault}{\sfdefault}\normalfont{}
  \centering
  \includegraphics[width=.5\linewidth]{gwas-design.pdf}
  \caption{Experimental design for human-nematode phenolog identification.
           We used GWAS candidates from the EBI-NHGRI catalog to identify
           disease associated genes, then used DIOPT to identify candidate
           orthologs in \cel{}. Orthologs were used to run phenotype, gene
           and tissue ontology enrichment analyses to identify disease
           phenologs.}
\label{fig:gwas}
\end{figure}

@@

\section*{Results}
\subsection*{Developing the WormBase enrichment suite}
% TODO add numbers instead of XXX
We developed the dictionaries for PEA and GEA using the same procedure as was
used for TEA~\cite{Angeles-Albores2016}. We generated a dictionary that included
terms with at least 50 annotating genes or more and had a similarity threshold
of 0.95 for PEA (the total number of terms in the dictionary was 251, annotated
by 9,169 genes for the version WS256);\@ and we generated a dictionary that
included terms with at least 50 annotating genes or more and had a similarity
threshold of 0.95 for GEA (the total number of terms in this dictionary is 271,
annotated by 14,636 genes for the version WS256). \@ Next, we benchmarked the
dictionaries on the same gene sets as TEA and obtained enrichment of all the
expected categories~\cite{Gaudet2004a, Spencer2011, Cinar2005, Watson2008a,
Pauli2006, Portman2004, Fox2007, Smith2010}. For example, on a gene set enriched
for embryonic muscle genes~\cite{Watson2008a}, the top two enriched phenotype
terms by q-value were `muscle system morphology variant' and `body wall muscle
thick filament variant'; the top two enriched GO terms were `myofibril' and
`striated muscle dense body'. For all the benchmarking results, see
supplementary information. Having generated and validated our dictionaries, we
proceeded to identify phenologs for several common human diseases.

\subsection*{Applying the WormBase enrichment suite}
To discover phenologs, we first needed to identify genes that contribute to a
disease in an unbiased manner. One way to discover gene associations in an
unbiased manner is to perform GWSA in human populations. Therefore, we used the
GWAS NHGRI-EBI Catalog~\cite{MacArthur2016} to identify genes associated with
human diseases. We found the best nematode candidate orthologs for these genes
using DIOPT~\cite{Hu2011} and applied our enrichment suite to each of these gene
regulatory networks.

\subsection*{Systemic lupus erythematosus}
Systemic lupus is an autoimmune disease that is believed to be polygenic in
nature~\cite{Mohan2015}. It mainly affects women and is characterized by painful
and swollen joints, hair loss, and fatigue~\cite{Lisnevskaia2014}. Since worms
do not have a cellular immune system, we were interested in what phenologs
corresponded to this disorder in \cel{}. To establish phenolog candidates, we
obtained \hlupus{} genes associated with the disease via GWAS studies, and found
\wlupus{} homolog candidates in \cel{}.

@@

TEA suggested that the `excretory duct cell' (5 genes, \qval{2}) and the
`posterior gonad arm' are overrepresented in this dataset. We also found that
the Pn.p cells P3.p through P8.p were enriched in this dataset (5 genes,
\qval{1}). GO enrichment pointed at `modification-dependent macromolecule
catabolic process' (23 genes, \qval{15}) as a molecular function that
characterizes this dataset. However, this GO term was enriched only due to a
single gene family, the \emph{skr} gene family. Almost the entire \emph{skr}
family was considered a candidate homolog to the SKP1 human gene, making the GO
enrichment suspect.

Enrichment of the terms for `aneuploidy', `meiotic chromosome segregation', and
`excess intestinal cells' were largely driven by the same gene group, which
includes \emph{cki-1}, and several \emph{skr} genes. On the other hand, `cell
fate transformation' and `male tail morphology' reflected the involvement of
developmental genes \emph{let-23}, and \emph{lin-12} among others. The term
`nonsense mRNA accumulation' was the result of \emph{pept-3}, \emph{smg-7},
\emph{tsr-1}, \emph{dhcr-7} and \emph{F08B4.7}. Therefore, we conclude that
systemic lupus erythematosus is potentially represented by a combination of
three phenotypes in \cel{}: A cell proliferation phenotype (either increased or
decreased), probably marked by increased aneuploidy; a developmental phenotype
involving cell fate transformation and leading to dysmorphias; and a molecular
phenotype involving impairment of the nonsense-mediated decay pathway. % TODO:
Is the text below true??? The results from the tissue enrichment analysis
highlighted three tissues that are particularly sensitive to \emph{lin}
mutations (the gonad, the excretory duct cell and the vulval precursor cells),
and the gonad arms undergo large quantities of nuclear proliferation.

\subsection*{Rheumatoid arthritis}
Rheumatoid arthritis is an auto-immune disease that is characterized by swollen
and painful joints that progressively deteriorate~\cite{Smolen2016}. Unlike
lupus, rheumatoid arthritis is not life-threatening, and comorbidity between
rheumatoid arthritis and lupus has not been described in past comorbidity
studies~\cite{Dougados2013}, suggesting that they may have at least partially
distinct genetic causes. We found \harthritis{} genes associated with rheumatoid
arthritis, for which we found \warthritis{} worm homolog candidates.

% TODO: Is no tissue enrichment observed bc poor annotations, or because
% all tissues randomly represented?
The only phenotype that was enriched for these orthologs was `short' (10 genes,
\qval{4}), even though 64 orthologs were associated with at least one phenotype
term. No tissue was enriched in this dataset. Because 82 genes are annotated to
have expression in at least one tissue, the lack of enrichment does not reflect
ignorance about the sites of expression of these genes. GEA showed that enriched
molecular functions for these genes include `collagen trimer' (22 genes,
\qval{15}). However, this term was enriched as the result of degeneracy in the
homolog candidates for the SFTPD gene. Other terms included `glycosylation' (10
genes, \qval{4}) and `Golgi apparatus' (11 genes, \qval{3}), but these terms
were enriched as the result of degenerate homolog candidates for the human gene
B3GNT7 which encodes a beta-1,3-N-acetylgalactosaminyltransferase.

@@

\subsubsection*{Enrichment in the ciliary neuronal transcriptome}
As an example of how ontology enrichment can improve our understanding of
transcriptomes, we selected a ciliary neuron dataset~\cite{Wang2015} and ran the
complete WormBase Enrichment Suite on it. Ciliary neurons are present in the
\cel{} male tail, but they are also present in the male cephalic sensillum and
hermaphrodites also have ciliated neurons. PEA reveals that the ciliated neuron
transcriptome is enriched for genes that are typically associated with `meiotic
chromosome segregation' (46 genes, \qval{5}), `aneuploidy' (42 genes, \qval{5})
and `spindle defective early embryos' (45 genes, \qval{2}) (see
Fig.~\ref{fig:cilia}).

@@

Although one interpretation of the results would be that microtubule genes are
driving the enrichment of these terms, another possibility is that there are
cell-cycle genes that are driving the enrichment of these phenotypes and
tissues. Indeed, GO enrichment shows terms such as `DNA replication' (29 genes,
\qval{5}), and `purine NTP-dependent helicase activity' (15 genes, \qval{1}).
Visual inspection of list in question reveals that cell-cycle and DNA
replication/repair genes are abundant in this transcriptome and include genes
such as \emph{atm-1}, \emph{dna-2}, or \emph{hpr-17}. This analysis reveals that
the ciliary neuron transcriptome is enriched in genes associated with
microtubules, but includes genes that are thought to interact with DNA either
via repair mechanisms or cell-cycle
control~\cite{Hofmann2000,Lee2003a,Bailly2010}.

>>>>

\usepackage[utf8x]{inputenc}
\usepackage{nameref}
\usepackage[right]{lineno}
\usepackage{amsmath}
\usepackage{booktabs}
\usepackage[numbers,super]{natbib}
\usepackage{changepage}
\usepackage{listings}
\usepackage{color}

@@

\definecolor{codegreen}{rgb}{0,0.6,0}
\definecolor{codegray}{rgb}{0.5,0.5,0.5}
\definecolor{codepurple}{rgb}{0.58,0,0.82}
\definecolor{backcolour}{rgb}{0.95,0.95,0.92}

\lstdefinestyle{mystyle}{
    backgroundcolor=\color{backcolour},
    commentstyle=\color{codegreen},
    keywordstyle=\color{magenta},
    numberstyle=\tiny\color{codegray},
    stringstyle=\color{codepurple},
    basicstyle=\footnotesize,
    breakatwhitespace=false,
    breaklines=true,
    captionpos=b,
    keepspaces=true,
    numbers=left,
    numbersep=5pt,
    showspaces=false,
    showstringspaces=false,
    showtabs=false,
    tabsize=2
}

\lstset{style=mystyle}

@@

\author[1,2]{David Angeles-Albores}
\author[1]{Raymond Y. N. Lee}
\author[1]{Juancarlos Chan}
\author[1,2,*]{Paul W. Sternberg}
\affil[1]{Division of Biology and Biological Engineering, Caltech,
Pasadena, CA, 91125, USA}
\affil[2]{Howard Hughes Medical Institute, Caltech, Pasadena, CA, 91125, USA}
\affil[*]{Corresponding author. Contact: pws@caltech.edu}
\renewcommand\Affilfont{\itshape\small{}}

@@

% \textbf{1} Department of Biology and Biological Engineering, Pasadena, CA,
% 91125, USA\\
% \textbf{2} Howard Hughes Medical Institute, Caltech, Pasadena, CA, 91125, USA
% \textbf{*} Corresponding author. Contact:pws@caltech.edu

\section*{ }
\textbf{Genome-wide experiments have the capacity to generate massive amounts of unbiased
data about an organism. To interpret these data, dimensionality reduction
techniques are required. One approach is to annotate genes using unified,
unambiguous vocabularies and to test experimental datasets for term enrichment
using probabilistic methods.
Although gene, phenotype and anatomy ontologies exist for \cel{}, no one
software offers enrichment analyses of all three ontologies.
Here, we present the WormBase Enrichment Suite, which offers users the
ability to test all nematode ontologies in parallel. We show that
the WormBase Enrichment Suite provides valuable insight into different
biological problems, and that phenotype enrichment analysis (PEA) can
help researchers identify disease phenologs, phenotypes that arise from the
action of orthologous genes across species, which can inform disease modeling in
\cel{}. The WormBase Enrichment Suite analysis can also shed light on
RNA-seq datasets by showing what molecular functions are enriched, which
phenotypes these functions are implicated in and what tissues are
overrepresented in the dataset. Finally, we explore the phenotype-anatomy
relationship, showing that a small subset of highly specific tissues are
disproportionately likely to affect the Egl phenotype, but inferring gene tissue
expression from an egg-laying (Egl) phenotype is limited to the largest tissues.
}

@@

\section*{Introduction}
The last decade has seen an explosion of techniques capable of genome-wide
measurements. Some examples of genome-wide tools include
RNA-seq~\cite{Mortazavi2008} to measure gene expression or
CHIP-seq~\cite{Johnson2007} to measure protein binding to chromatin. These tools
are capable of generating large quantities of data. Understanding these data,
and generating hypotheses from them remains challenging. A common approach used
to understand these datasets is to reduce the dimensionality of the data via
enrichment analyses of ontologies~\cite{TheGeneOntologyConsortium2000a}, which
helps researchers understand what terms are overrepresented beyond random
levels. By analyzing overrepresented terms in aggregate, researchers can better
understand what biological processes were most affected in a given experiment,
and form hypotheses about what is happening~\cite{Rhee2008}. This approach is
limited by what ontologies can be tested for enrichment. The best-known ontologies
for biological research are the Gene Ontologies (GO), which provide a controlled
language to describe the biological, cellular and molecular aspects and functions of
genes~\cite{TheGeneOntologyConsortium2000a}. In \cel{}, anatomy and
phenotype ontologies exist with which to describe \cel{} anatomy and phenotypes
respectively~\cite{Schindelman2011,Lee2003}. These latter ontologies are developed by
professional curators at WormBase, which is a repository of \cel{}
data~\cite{Howe2016}. However, enrichment tools currently only exist for gene and tissue
ontologies in the community today (see for
example~\cite{Chikina2009,Mi2013,Angeles-Albores2016}). Another limitation is
that tissue enrichment testing is not offered on the same websites as GO
enrichment testing, which requires users to test their data on different
websites that use different methodologies and databases to detect enrichment.

Another use of enrichment tools is for evolutionary comparison purposes.
In molecular biology it is often useful to know when a gene is orthologous
between two species---that is to say, common by descent---because knowledge of
orthology often brings with it knowledge of function. Indeed, many
important gene regulatory networks (GRNs) are conserved between organisms as
highly diverged as nematodes and humans (for example, see~\cite{Sternberg1998}).
While genes and GRNs may be conserved between species, their outputs often differ.
For example, the gene Pax6 (Eyeless in \emph{Drosophila melanogaster})
is involved in eye formation in humans and fruit-flies~\cite{Quiring1994}. Although
nematodes have conserved this gene, they do not have eyes~\cite{Zhang1995,Chisholm1995}.
The concept of a phenolog has been put forward to explain relationships between
phenotypes that have the same underlying genetic
regulatory network~\cite{McGary2010,Lehner2013}.
% Formally, two phenotypes are phenologs of each other if the orthologs of the
% genes that cause a phenotype in an organism cause a second phenotype in another.

To study a clinically relevant disease in a non-human, an appropriate model has
to be established. One method to establish a disease model
in \cel{} is to link a disease to a causal gene, then to identify the
\cel{} ortholog, study the function of the ortholog in order to
extrapolate back to humans. However, this method relies on the existence of
known disease genes and requires that the ortholog have a phenotype that can be
reliably identified and studied. A fundamentally different way to establish a
disease model in \cel{} would be to identify the phenologs of the disease to be
studied in \cel{} by identifying disease-associated human genes in an unbiased
manner through genome-wide association studies (GWAS) and identified candidate
ortholog genes in \cel{}. The orthologs can be used to identify \cel{} disease
phenologs, which can in turn be used as the basis for screens to identify genes
that are associated with that phenolog. Approaches similar to this have been
successfully used in the past to make non-obvious links between phenotypes in
different species~\cite{McGary2010}.

The concept of a phenolog can also be useful when applied within a species. In
\cel{}, not all phenotypes are equally easy to study. Although genome-wide
measurements can help elucidate the genetic network underlying a phenotype,
devising screens to test which genes are functionally important can be
difficult. A common strategy to study phenotypes that are difficult to screen is
to select an easier-to-screen phenolog, and to test positive hits for the original
phenotype of interest afterwards. For example, candidate genes involved in \cel{}
aging control can be more easily identified using stress
assays~\cite{Kim2007a,Mehta2009}. Currently, which phenotypes are screened for
depends on the researcher's prior experience. By formalizing phenotype enrichment
analysis as a tool with which to analyze gene sets, researchers should be able
to formally establish phenologs, which has consequences for screen design.

An additional problem with genome-wide queries of \cel{} states (be they
developmental, such as L1, L2, dauer; behavioral states such as awake versus
asleep; or other) is that they do not always have a straightforward
interpretation in terms of phenotypes. In these situations, researchers must
rely on intuition to select a phenotype for which to screen. As a result, many
loci may go unexplored that would prove relevant. The question of how to design
a screen that is maximally informative is an important question that has so far
not been addressed within this community.

To facilitate understanding of large datasets, we have developed the WormBase
Enrichment Suite (URL:\
\url{http://www.wormbase.org/tools/enrichment/tea/tea.cgi}), which allows users to
rapidly perform phenotype, tissue and gene ontology enrichment analyses (PEA,
TEA and GEA respectively) using a unified methodology. As examples of use cases,
we applied our tools towards
the unbiased discovery of phenologs of multigenic, complex diseases including
systemic lupus erythematosus, obesity and obesity-related traits, and rheumatoid
arthritis by using genes associated with these diseases via genome-wide
association studies. We illustrate the utility of the complete enrichment suite
for finding new relationships in complex data by analyzing a ciliary neuron
transcriptome~\cite{Wang2015} as a test case. Finally, we show that the
dictionaries generated for these enrichment analyses can help elucidate the
contributions of specific tissues to specific phenotypes.

@@

Enrichment analyses are performed as described in Angeles-Albores
\emph{et al}~\cite{Angeles-Albores2016}.
Briefly, a static dictionary is constructed for each ontology using the most
recent WormBase release annotation information. This dictionary is generated
using a previously described trimming algorithm~\cite{Angeles-Albores2016} to remove
terms that are too general and uninformative as well as terms that are very
specific but do not have many annotations. When the user provides a set of genes
to test against, we use a hypergeometric model to test whether a given ontology
term is occurring at a greater frequency than expected by random chance.

\subsection*{Human disease phenolog identification}
We used the GWAS EBI-NHGRI catalog~\cite{MacArthur2016} to extract information
on all genome-wide association studies deposited there. We only selected traits
that had $>300$ associated genes. We identified 24 traits that met our criteria.
Next, we used DIOPT~\cite{Hu2011} to identify candidate \cel{} orthologs for the genes
associated with these traits. Briefly, DIOPT combines a large number of methods
for identifying orthologs and returns ortholog candidates associated with a
compound score. Depending on the score, orthologs can be considered `high',
`moderate' or `low' rank, reflecting confidence in the orthology. Many-to-one and
one-to-many orthology relationships are allowed in DIOPT, reflecting a mixture of
uncertainty and family expansion/reduction. For our study, we only accepted
ortholog candidates with `high' or `moderate' scores and we did not insist on a
one-to-one relationship between genes.

\begin{figure}[htbp]
  \renewcommand{\familydefault}{\sfdefault}\normalfont{}
  \centering
  \includegraphics[width=.5\linewidth]{gwas-design.pdf}
  \caption{Experimental design for human-nematode phenolog identification.
           We used GWAS candidates from the EBI-NHGRI catalog to identify
           disease associated genes, then used DIOPT to identify candidate
           orthologs in \cel{}. Orthologs were used to run tissue, phenotype,
           gene ontology enrichment analyses to identify disease
           phenologs.}
\label{fig:gwas}
\end{figure}

@@

\section*{Results}
\subsection*{Developing the WormBase enrichment suite}
% TODO add numbers instead of XXX
We developed the dictionaries for PEA and GEA using the same procedure as was
used for TEA~\cite{Angeles-Albores2016}. We generated a dictionary that included
terms which were annotated to at least 50 genes or more and had a similarity threshold
of 0.95 for PEA (the total number of terms in the dictionary was 251, annotated
to 9,169 genes for the version WS256);\@ and we generated a dictionary that
included terms which are annotated to at least 50 genes or more and had a similarity
threshold of 0.95 for GEA (the total number of terms in this dictionary is 271,
annotated to 14,636 genes for the WormBase version WS256). \@ Next, we benchmarked the
dictionaries on the same training gene sets as TEA and obtained enrichment of all the
expected categories~\cite{Gaudet2004a, Spencer2011, Cinar2005, Watson2008a,
Pauli2006, Portman2004, Fox2007, Smith2010}. For example, on a gene set enriched
for embryonic muscle genes~\cite{Watson2008a}, the top two enriched phenotype
terms by q-value were `muscle system morphology variant' and `body wall muscle
thick filament variant'; the top two enriched GO terms were `myofibril' and
`striated muscle dense body'. Having generated and validated our dictionaries,
we proceeded to identify phenologs for several common human diseases.

\subsection*{Applying the WormBase enrichment suite}
To discover phenologs, we first needed to identify genes that contribute to a
disease in an unbiased manner. One way to discover gene associations in an
unbiased manner is to perform GWAS in human populations. Therefore, we used the
GWAS NHGRI-EBI Catalog~\cite{MacArthur2016} to identify genes associated with
human diseases. We found the best nematode candidate orthologs for these genes
using DIOPT~\cite{Hu2011} and applied our enrichment suite to each of these gene
regulatory networks.

\subsection*{Systemic lupus erythematosus}
Systemic lupus is an autoimmune disease that is believed to be polygenic in
nature~\cite{Mohan2015}. It mainly affects women and is characterized by painful
and swollen joints, hair loss, and fatigue~\cite{Lisnevskaia2014}. Since worms
do not have a cellular immune system, we were interested in what phenologs
corresponded to this disorder in \cel{}. To establish phenolog candidates, we
obtained \hlupus{} genes associated with the disease via GWAS studies, and found
\wlupus{} ortholog candidates in \cel{}.

@@

TEA suggested that the `excretory duct cell' (5 genes, \qval{2}) and the
`posterior gonad arm' are overrepresented in this dataset. We also found that
the Pn.p cells P3.p through P8.p were enriched in this dataset (5 genes,
\qval{1}). GO enrichment pointed to `modification-dependent macromolecule
catabolic process' (23 genes, \qval{15}) as a molecular function that
characterizes this dataset. However, this GO term was enriched only due to a
single gene family, the \emph{skr} gene family. Almost the entire \emph{skr}
family was considered a candidate ortholog to the SKP1 human gene, making the GO
enrichment suspect.

Enrichment of the terms `aneuploidy', `meiotic chromosome segregation', and
`excess intestinal cells' were largely driven by the same gene group, which
includes \emph{cki-1}, and several \emph{skr} genes. On the other hand, `cell
fate transformation' and `male tail morphology' reflected the involvement of
developmental genes \emph{let-23}, and \emph{lin-12} among others. The term
`nonsense mRNA accumulation' was the result of \emph{pept-3}, \emph{smg-7},
\emph{tsr-1}, \emph{dhcr-7} and \emph{F08B4.7}. Therefore, we conclude that
systemic lupus erythematosus is potentially represented by a combination of
three phenotypes in \cel{}: A cell proliferation phenotype (either increased or
decreased), probably marked by increased aneuploidy; a developmental phenotype
involving cell fate transformation and leading to dysmorphias; and a molecular
phenotype involving impairment of the nonsense-mediated decay pathway.
The results from the tissue enrichment analysis
highlighted three tissues that are particularly sensitive to \emph{lin}
mutations (the gonad, the excretory duct cell and the vulval precursor cells).
Gonad development also requires large quantities of nuclear proliferation.

\subsection*{Rheumatoid arthritis}
Rheumatoid arthritis is an auto-immune disease that is characterized by swollen
and painful joints that progressively deteriorate~\cite{Smolen2016}. Unlike
lupus, rheumatoid arthritis is not life-threatening, and comorbidity between
rheumatoid arthritis and lupus has not been described in past comorbidity
studies~\cite{Dougados2013}, suggesting that they may have at least partially
distinct genetic causes. We found \harthritis{} genes associated with rheumatoid
arthritis, for which we found \warthritis{} worm ortholog candidates.

% TODO: Is no tissue enrichment observed bc poor annotations, or because
% all tissues randomly represented?
The only phenotype that was enriched for these orthologs was `short' (10 genes,
\qval{4}), even though 64 orthologs were associated with at least one phenotype
term. No tissue was enriched in this dataset. Because 82 genes are annotated to
have expression in at least one tissue, the lack of enrichment does not reflect
ignorance about the sites of expression of these genes. GEA showed that enriched
molecular functions for these genes include `collagen trimer' (22 genes,
\qval{15}). However, this term was enriched as the result of degeneracy in the
ortholog candidates for the SFTPD gene. Other terms included `glycosylation' (10
genes, \qval{4}) and `Golgi apparatus' (11 genes, \qval{3}), but these terms
were enriched as the result of degenerate ortholog candidates for the human gene
B3GNT7 which encodes a beta-1,3-N-acetylgalactosaminyltransferase.

@@

\subsubsection*{Enrichment in the ciliary neuronal transcriptome}
As an example of how ontology enrichment can improve our understanding of
transcriptomes, we selected a ciliary neuron dataset~\cite{Wang2015} and ran the
complete WormBase Enrichment Suite on it (see Listing~\ref{code} for a brief
tutorial on how to do this in Python). Ciliary neurons are present in the
\cel{} male tail, but they are also present in the male cephalic sensillum and
hermaphrodites also have ciliated neurons. PEA reveals that the ciliated neuron
transcriptome is enriched for genes that are typically associated with `meiotic
chromosome segregation' (46 genes, \qval{5}), `aneuploidy' (42 genes, \qval{5})
and `spindle defective early embryos' (45 genes, \qval{2}) (see
Fig.~\ref{fig:cilia}).

\begin{lstlisting}[language=Python, caption=Brief programmatic example of how
to perform a complete phentoype enrichment analysis in Python., label={code}]
  import pandas as pd
  import tissue_enrichment_analysis as tea
  import matplotlib.pyplot as plt

  # read in the file to be analyzed
  # this file has two columns of interest:
  # `gene_name', contains WormBase IDs
  # `pval', which contains the p-value of each ORF
  df = pd.read_csv('path/to/data_file.csv')

  # import the phenotype dictionary to load into tea
  phenotype_dictionary = tea.fetch('phenotype')

  # get the gene IDs to be analyzed
  # we only want those genes with pval < 0.05
  gene_names = df[df.pval < 0.05].gene_IDs

  # run the enrichment analysis.
  # the line below generates a new pandas dataframe with the results
  results = tea.enrichment_analysis(wbid_list, tissue_df=phenotype_dictionary)

  # save the results to a file:
  results.to_csv('path/to/analysis_results.csv', index=False)

  # plot the results
  # The `kind' variable depends on the enrichment analysis:
  # Tissue, Phenotype or GO.
  tea.plot_enrichment_results(results, kind='Phenotype')

  # save the plot
  plt.savefig('path/to/graph.pdf')
\end{lstlisting}

@@

Although one interpretation of the results would be that microtubule genes are
driving the enrichment of these terms, another possibility is that there are
cell-cycle genes that are driving the enrichment of these phenotypes and
tissues. Indeed, GO enrichment shows terms such as `DNA replication' (29 genes,
\qval{5}), and `purine NTP-dependent helicase activity' (15 genes, \qval{1}).
Visual inspection the genes driving enrichment of these terms reveals that
cell-cycle and DNA replication/repair genes are abundant in this transcriptome
and include genes such as \emph{atm-1}, \emph{dna-2}, or \emph{hpr-17}. This
analysis reveals that the ciliary neuron transcriptome is enriched in genes
associated with microtubules, but includes genes that are thought to interact
with DNA either via repair mechanisms or cell-cycle
control~\cite{Hofmann2000,Lee2003a,Bailly2010}.

\subsection*{Deconstructing phenotype-tissue relationships}
\subsubsection*{Tissue enrichment on the Egl gene set reveals cellular
components of the phenotype}
How does a phenotype emerge? With the tools that we have
developed, it is possible to understand what tissues contribute to a phenotype
in a probabilistic framework. In other words, we can extract all genes
associated with a particular phenotype, then search for tissue terms that are
enriched to understand how a phenotype arises from interactions between
anatomical regions. As a test of this, we selected the egg-laying defective
(Egl) phenotype. In \cel{}, egg-laying is a complex behavior that involves a
large number of tissues~\cite{Li1990}. The somatic gonad acts as a repository
for the eggs, the uterine seam cells help protect the uterus, and a variety of
muscles help contract the uterus and open the vulva to lay an
egg~\cite{Sulston1977}. The vulva must be well-formed to allow passage of an
egg, and the hermaphrodite-specific neuron (HSN) is involved in the control of
egg-laying~\cite{Schafer2005}. The complexity of the interactions required for
egg-laying makes understanding the Egl phenotype in terms of tissue interactions
a challenging task.
